---
title: "Linear Models Version1 - Ramsay"
author: "Juan Diego Ariza Sanchez"
date: "2026-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(fda)
data("daily")
```

# Functional Linear Models

This is a Markdown that implements the most basic models explained in the book of Ramsay, the codes in here can be used for future reference. Some mathematical notation will also be deployed.

## Functional Linear Models for Scalar Responses

Similar as the multivariate linear model but exploiting the mesh of times one have the following functional linear model for scalar response: $$
y_i = \alpha_0 + \int x_i(t)\beta (t) dt + \epsilon_i
$$ How to estimate $\beta(t)$? There are two options:

-   Use a basis coefficient expansion $\beta(t) = \sum_k^K c_k \phi_k(t)$
-   Use PCA

### Example

Here we model the logarithm of annual precipitation for 35 Canadian weather stations from their temperature profiles. In other words, $y_i$ is the log annual precipitation and $x_i(t)$ are the temperature profiles. Now, lets create the scalar variable (annualprec) and the functional one (tempfd)

```{r Functional preprocess}
# Scalar variable
annualprec = log10(apply(daily$precav,2,sum))
# Functional variable
tempbasis =create.fourier.basis(c(0,365),65)
tempSmooth=smooth.basis(day.5,daily$tempav,tempbasis)
tempfd =tempSmooth$fd
```

For educational purpouses, let's plot the functional variable

```{r}
plot(tempfd)
```

And now, let's create the regressor structure

```{r X(t)}
templist = vector("list",2)
templist[[1]] = rep(1,35)
templist[[2]] = tempfd
```

#### First strategy to estimate $\beta(t)$

Keep the dimensionality $K$ of $b$ small relative to $N$. Here let's set $K=5$.

```{r Basis Coefficient 1}
conbasis = create.constant.basis(c(0,365))
betabasis = create.fourier.basis(c(0,365),5)
betalist = vector("list",2)
betalist[[1]] = conbasis
betalist[[2]] = betabasis
```

Now, we run the function *fRegress* 

```{r Model 1}
fRegressList = fRegress(annualprec,templist,betalist)
```

The result *betaestlist* contains the estimated regression coefficient functions. Each of these is a functional parameter object. We can plot the estimate of the regression function for the temperature profiles:

```{r}
# Intercept
coef(betaestlist[[1]])
#Beta (t)
betaestlist = fRegressList$betaestlist
tempbetafd = betaestlist[[2]]$fd
plot(tempbetafd, xlab="Day", ylab="Beta for temperature")
```
##### Quality of the strategy 1
Let's assess the quality of the previous fit, we can compute the error sums of squares

```{r quality 1}
# Estimated y (\hat(y))
annualprechat1 = fRegressList$yhatfdobj
# Errors
annualprecres1 = annualprec - annualprechat1
# SES
SSE1.1 = sum(annualprecres1^2) #Our model
SSE0 = sum((annualprec - mean(annualprec))^2) #Null model
```

We can now compute the squared multiple correlation and the usual F-ratio for comparing these two fits.

```{r}
(RSQ1 = (SSE0-SSE1.1)/SSE0)
(Fratio1 = ((SSE0-SSE1.1)/5)/(SSE1.1/29))
```

#### Second strategy to estimate $\beta(t)$
Here we will include a roughness penalty. Instead of using a low-dimension basis. The combination of a high-dimensional basis with a roughness penalty reduces the possibilities that either (a) important features are missed or (b) extraneous features are forced into the image by using a basis set that is too small for the application.
This is a more powerful approach!

